---
title: "The musicXML package"
subtitle: "GettingStarted"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,results='hide',
                      fig.path = "man/figures/README-")
```

# Introduction
[MusicXML](https://www.musicxml.com/) is a file format for representing musical scores ([Wikipedia](https://en.wikipedia.org/wiki/MusicXML)). It has become the standard format for sharing music notation in a digital form and is hence supported by [most musical notation softwares](https://www.musicxml.com/software/).


While MusicXML is easily understood by computers, it is less friendly to the human eye. As an illustration, consider [Wikipedia's example](https://en.wikipedia.org/wiki/MusicXML#Example): writing a simple whole note in MusicXML format looks tedious, and writing a whole score by hand would certainly take some patience.

The `musicXML` package allows using R to write files in MusicXML format. It is built as a minimalistic interface to the MusicXML format: the basic musical objects (notes, measures etc.) of a score can be defined in R and the package will take care of translating them into a proper MusicXML format. The resulting file can then be opened and further modified using your favorite musical score software (e.g. the free [MuseScore](https://musescore.org) or the simple online viewer/player [SoundSlice](https://www.soundslice.com/musicxml-viewer/)). One particular application is data sonification, as will be illustrated in this vignette

So let's load the package and let's get it started!

```{r}
library(musicXML)
```

# Manipulating basic objects: notes, measures and scores

The `score` to be written may contain several `measures`, each of which may contain several `notes`. These can be defined as described below.

## Notes

A `note` is defined by the following three attributes:

1. its `pitch`, for instance 'C5', 'Gb4', 'G#6', etc. (using the [scientific pitch notation](https://en.wikipedia.org/wiki/Scientific_pitch_notation))

2. its `duration`, for instance a whole note, a dotted half, a triplet quarter, etc.

3. its `loudness`, controlling the volume at which it is played.

Let's say we want to define a whole-note C5 played *forte* for instance. We start by defining its `pitch` using a simple character string comprising: (i) the base pitch: one letter in ABCDEFG; (ii) the alteration: '' (none), 'b' (flat) or '#' (sharp); (iii) the octave: an integer between 0 and 9.

```{r}
p <- pitch(string='C5') # defines a pitch object
```

We then define the `duration` using an integer with the following meaning: 1 = whole, 2 = half, 4 = quarter, 8 = eighth, etc. down to 64 = 64th.

```{r}
d <- duration(type=1) # defines a duration object
```

We then define the `loudness`, which is expressed as a number between 0 and 141, with the following guide: 37: *pp*, 54: *p*, 71: *mp*, 89: *mf*, 107: *f*, 124: *ff*.

```{r}
l <- 107 # defines loudness of the note
```

The note can finaly be defined by stitching pitch, duration and loudness together as follows: 

```{r}
n <- note(p,d,l) # defines a note object
```

Of course all this can be done in a single line of code:

```{r}
n <- note(p=pitch('C5'),d=duration(1),l=107)
```

If you want to have a look at what this note would look like in musicXML format, you can use the function `toMXL`. Note that the resulting string is not sufficient to define a proper musicXML file, it just corresponds to a "note" block of the future complete file.

```{r,results='markup'}
toMXL(n)
```

Let's define a few more notes to start a melody and to illustrate a couple of useful options:

```{r}
# a quarter G3 played fff
n1 <- note(p=pitch('G3'),d=duration(4),l=140)
# a dotted 8th D4 played f. Note the use of 'dot=TRUE' in duration. There is also a 'triplet=' option.
n2 <- note(p=pitch('D4'),d=duration(8,dot=TRUE),l=100)
# a 16th G4 played mp
n3 <- note(p=pitch('G4'),d=duration(16),l=80)
# two quarter Bb4 played p and tied together (tie2next=TRUE - you could also use tie2previous=TRUE on the second one) 
n4 <- note(p=pitch('Bb4'),d=duration(4),l=60,tie2next=TRUE)
n5 <- note(p=pitch('Bb4'),d=duration(4),l=60)
```

## Measures

The five notes defined above add up to 4 full beats; they can hence be placed into a single 4/4 `measure` as shown below:

```{r}
# defining a measure. 'number=' is the measure index in the whole score
m <- measure(number=1,notes=list(n1,n2,n3,n4,n5))
```

Alternatively, you could decide to rather use two 2/4 measures:

```{r}
# note the use of 'beats=' and 'beatType=' to modify the time signature
m1 <- measure(number=1,notes=list(n1,n2,n3),beats=2,beatType=4)
m2 <- measure(number=2,notes=list(n4,n5),beats=2,beatType=4)
```

It is also possible to change the key signature of a measure. For instance the five notes defined above would fit in a minor G scale, with 3 flats in the key signature:

```{r}
# note the use of 'keySignature=' to add flats (negative values) of sharps (positive values) to the key
m1 <- measure(number=1,notes=list(n1,n2,n3),beats=2,beatType=4,keySignature=-3,mode='minor')
m2 <- measure(number=2,notes=list(n4,n5),beats=2,beatType=4,keySignature=-3,mode='minor')
```

## Score
Finally, all measures can be grouped together to define a `score`:

```{r}
s <- score(list(m1,m2))
```

Note that it is possible to define a multi-part score as follows:

```{r}
part1 <- list(m1,m2)
part2 <- list(m2,m2)
s <- score(list(part1,part2))
```

The score can finally be written into a musicXML file. You can open it with a text editor to see what it looks like, but more interestingly, you can open it with a musical software to see, listen and modify the resulting musical score (for a quick online look try [SoundSlice](https://www.soundslice.com/musicxml-viewer/)). 

```{r}
writeMXL(s,file='myFirstScore.xml')
```

# Application to data sonification

## General principles 

Data [sonification](https://en.wikipedia.org/wiki/Sonification) refers to the transformation of data into sound, using some algorithmic process. This can be achieved in [many ways](https://sonification.de/handbook/), but here we focus on the approach known as [*parameter mapping*](https://sonification.de/handbook/chapters/chapter15/): the values taken by the data are mapped into some attributes (or parameters) of notes, typically their pitch or loudness, or possibly their duration. This is very much the same process as the mapping performed in data visualization, where data values are mapped into e.g. the color, size of type of symbols in a graph or map.

Consider, as an example, the Wagga Wagga dataset which comes with this package. It contains the annual precipitation and temperature time series in the city of Wagga Wagga, New South Wales, Australia, as plotted below. A possible sonification of this dataset is to map precipitation into pitch (so that wet years correspond to high-pitched notes) and temperature to loudness (so that warm years correspond to loud notes). Let's see how this can be done.

```{r}
par(mfrow=c(2,1))
plot(WaggaWagga$Year,WaggaWagga$Precipitation,type='l',xlab='Year',ylab='precip. [mm]')
plot(WaggaWagga$Year,WaggaWagga$Temperature,type='l',xlab='Year',ylab='temp. [C]')
```

## Sonification of the Wagga Wagga dataset 

We start with the mapping of temperature to loudness. This is achieved using the `loudnessMapping` function which takes a vector of data as input and returns a vector of corresponding loudnesses using a simple linear interpolation between a minimal and a maximal loudness. Although not detailled here, there are options to use a nonlinear mapping, see ?loudnessMapping.

```{r}
# Compute loudnesses from temperatures
llist <- loudnessMapping(WaggaWagga$Temperature,lMin=20,lMax=141)
# Check the relation is just a simple linear interpolation
plot(WaggaWagga$Temperature,llist,pch=19,col=rgb(0,0,0,0.2),xlab='Temp. [C]',ylab='loudness')
```

We now perform the mapping of precipitation to pitch. This is a bit trickier because instead of doing a continuous mapping, we want to map values to a discrete set of notes that 'sound good'. To achieve this, we need to define this set of notes, typically using a musical scale (a 2-octave pentatonic A scale in the example below). Note that the output of `pitchMapping` function is not a vector of numeric values, but rather a list of `pitch` objects.

```{r}
scale <- c("A3", "C4", "D4", "E4", "G4", "A4", "C5", "D5", "E5", "G5")
plist <- pitchMapping(WaggaWagga$Precipitation,pitches=scale)
```

We now have a list of loudnesses, a list of pitches, we can turn them into a list of notes as shown below. Note that by default the duration of each note will be a 16th note, but this can be changed, see ?getNotes.

```{r}
# Take a list of pitches/loudnesses and create a list of notes
notes <- getNotes(pitches=plist,loudnesses=llist)
```

We can now stack these notes into measures having a given time signature (4/4 is used below). Note that the function `getMeasures` automatically groups notes to ensure that each measure has the correct amount of beats.

```{r}
# Take a list of notes and stack them into a list of measures.
m <- getMeasures(notes=notes,beats=4,beatType=4)
```

Finally, we can now define the `score` and write it into a musicXML file. 

```{r}
s <- score(m)
writeMXL(s,file='WaggaWagga.xml')
```

That's it! The dataset has been transformed into a musical score. The job of the musicXML R package stops here, but of course this does not mean that the sonification process is necessarily over: the score can be further read, played and modified using a specialized music software. 

Note that data sonification is particularly interesting when it is combined with data animation, as shown in the example below based on the package [gganimate](https://gganimate.com/index.html). Note that the musicXML file has to be transformed into an audio format (such as .mp3) to achieve this. Unfortunately this is not yet implemented in this R package, but most specialized software can do it (I did it with [MuseScore](https://musescore.org)).

```{r,results='markup'}
library(tidyr);library(gganimate)
# Modify the shape of the WaggaWagga dataset to facilitate plotting
DF <- pivot_longer(WaggaWagga,-Year) # function from tidyr
# Plot precipitation and temperature time series using ggplot
g <- ggplot(DF,aes(x=Year,y=value))
g <- g + geom_line(aes(color=name),size=1)+geom_point(size=4)
g <- g + scale_color_manual(values = c('blue','red'),guide=FALSE)
g <- g + facet_wrap(vars(name),ncol=1,scales='free_y')
# Make it look nicer
g <- g+theme_bw()+theme(axis.title=element_text(size=18), 
          axis.text=element_text(size=14),
          strip.text=element_text(size=18))
# Create an animated plot
g <- g + transition_reveal(Year)
# 'Render' the animated plot into a .mp4 movie
bpm=120 # tempo (in beats per minute) used to create the audio file
bps=(bpm/60) # convert to beats per second
fps=bps*4 # convert to frames per second (there are 4 16th note per beat)
animate(g,nframes=NROW(WaggaWagga),fps=fps,width=1280,height=720,
        renderer = av_renderer('WaggaWaggaMelody.mp4',audio='vignettes/WaggaWagga.mp3'))
```

# Conclusion
The musicXML package has been built with the idea of providing a minimalistic R interface to the musicXML format. In particular, its intended usage is to transform a series of data into a musical score, by mapping data values into pitch, loudness or duration. This transformation is the first step of a data sonification endeavor, however it is certainly not the last. More advanced music-oriented modifications of this score can be performed (e.g. changing instrument, using effects, etc.), but this is deferred to specialized music software.

